---
title: "Class 11: Genomics"
author: "Suzanne Enos"
format: pdf
---
> Q5: What proportion of the Mexican Ancestry in Los Angeles sample population (MXL) are homozygous for the asthma associated SNP (G|G)? 

```{r}
gene <- read.csv("rs8067378.csv", row.names = 1)
head(gene)
```

```{r}
table(gene$Genotype..forward.strand.)
```

```{r}
table(gene$Genotype..forward.strand.)/nrow(gene)
```



# Population Scale Analysis

> Q13. Read this file into R and determine the sample size for each genotype and their
corresponding median expression levels for each of these genotypes.

```{r}
results <- read.table("rs8067378_ENSG00000172057.6.txt")
head(results)
```

```{r}
summary(results)
```

```{r}
table(results$geno)
```


```{r}
inds <- results$geno == "G/G"
summary(results$exp[inds])
```

```{r}
inds <- results$geno == "A/G"
summary(results$exp[inds])
```

```{r}
inds <- results$geno == "A/A"
summary(results$exp[inds])
```

>Q14. Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?

```{r}
library(ggplot2)

ggplot(results) + 
  aes(geno, exp, col = geno) +
  geom_boxplot(notch = T) +
  geom_jitter(alpha = 0.3, width = 0.3)
```

