---
title: "Class 17: Vaccination Rate Mini Project"
author: "Suzanne Enos"
format: gfm
---
# Background

We're exploring a dataset on statewide vaccination rate from CA.gov

The goal of this hands-on mini-project is to examine and compare the Covid-19 vaccination rates around San Diego.

We will start by downloading the most recently dated “Statewide COVID-19 Vaccines Administered by ZIP Code” CSV file from: https://data.ca.gov/dataset/covid-19-vaccine-progress-dashboard-data-by-zip-code

## Data Input

```{r}
vax <- read.csv("covid19vaccinesbyzipcode_test.csv")
head(vax)
```

Correct way to store numeric dates: year-month-day

>Q1. What column details the total number of people fully vaccinated?

persons_fully_vaccinated

>Q2. What column details the Zip code tabulation area? 

zip_code_tabulation_area

> What is the earliest date in this dataset?

2021-01-05

>Q4. What is the latest date in this dataset?

```{r}
tail(vax)
```
```{r}
vax$as_of_date[nrow(vax)]
```

2023-03-07

# Use Skimr to get an overview of the dataset

If you just want to use one function from a package and not load the whole package: package_name:: instead of library() 

```{r}
skimr::skim(vax)
```
>Q5. How many numeric columns are in this dataset? 

13

>Q6. Note that there are “missing values” in the dataset. How many NA values there in the persons_fully_vaccinated column? 

16621
```{r}
n.missing <- sum(is.na(vax$persons_fully_vaccinated))
n.missing
```


>Q7. What percent of persons_fully_vaccinated values are missing (to 2 significant figures)? 

```{r}
round(n.missing/nrow(vax)*100, 2)

```

# Working with dates

We will use the lubridate package to help ease the pain of working with times and dates
```{r}
library(lubridate)
```

```{r}
today()
```

```{r}
today() - ymd(vax$as_of_date[1])
```

How old am I in days?

```{r}
today() - ymd("2000-01-27")
```

>Q9. How many days have passed since the last update of the dataset?

```{r}
today() - ymd(vax$as_of_date[nrow(vax)])
```

```{r}
vax$as_of_date <- ymd(vax$as_of_date)
```

```{r}
today() - vax$as_of_date[nrow(vax)]
```

>Q10. How many unique dates are in the dataset (i.e. how many different dates are detailed)? 

```{r}
length(unique(vax$as_of_date))
```

```{r}
library(dplyr)

n_distinct(vax$as_of_date)
```

# Working with zipcodes

Can use zipcodeR package to make this easier to work with

```{r}
library(zipcodeR)
geocode_zip("30022")
```

```{r}
zip_distance('92037', '92109')
```
```{r}
reverse_zipcode(c('92037', '92109', '30605') )
```

> Finding best and worst ratio of household income to home prices

Find all SD zipcodes
```{r}
sd.zip <- unique(vax$zip_code_tabulation_area[vax$county == "San Diego"])
length(sd.zip)
```

```{r}
sd.eco <- reverse_zipcode(sd.zip)
sd.eco
```

Most expensive area?
```{r}
ord <- order(sd.eco$median_household_income, decreasing=T)
head(sd.eco[ord,])
```

```{r}
head(arrange(sd.eco, desc(median_home_value)))
```


```{r}
sd.eco$median_home_value
```

# Focus on the San Diego area

Using **dplyr** here
```{r}
sd <- filter(vax, county == "San Diego")

nrow(sd)
```

>Q11. How many distinct zip codes are listed for San Diego County? 

```{r}
n_distinct(sd$zip_code_tabulation_area)
```

>Q12. What San Diego County Zip code area has the largest 12 + Population in this dataset?

```{r}
ind <- which.max(sd$age12_plus_population)
sd[ind,]
```
>Q13. What is the overall average “Percent of Population Fully Vaccinated” value for all San Diego “County” as of “2023-03-07”? 

```{r}
library(dplyr)
sd.latest <- filter(sd, as_of_date == "2023-03-07")
mean(sd.latest$percent_of_population_fully_vaccinated, na.rm = TRUE)
```

>Q14. Using either ggplot or base R graphics make a summary figure that shows the distribution of Percent of Population Fully Vaccinated values as of “2023-02-28”?

```{r}
library(ggplot2)
ggplot(sd.latest) +
  aes(sd.latest$percent_of_population_fully_vaccinated) +
  geom_histogram(bins = 20) +
  theme_bw() +
  xlab("Percent of People Vaccinated")
```

# Focus on UCSD/La Jolla

```{r}
ucsd <- filter(sd, zip_code_tabulation_area == "92037")
ucsd[1,]$age5_plus_population
```

```{r}
ljplot <- ggplot(ucsd) +
  aes(x = as_of_date, y = percent_of_population_fully_vaccinated) +
  geom_point() +
  geom_line(group=1) +
  ylim(c(0,1)) +
  labs(title = "Vaccination rate for La Jolla CA 92037", x = "Date", y = "Percent Vaccinated") +
  theme_bw()
ljplot
```

# Comparing to similar sized areas

```{r}
vax.36 <- filter(vax, age5_plus_population > 36144 &
                as_of_date == "2023-03-07")
head(vax.36)
```

How many unique zipcodes?
```{r}
n_distinct(vax.36$zip_code_tabulation_area)
```

>Q16. Calculate the mean “Percent of Population Fully Vaccinated” for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2023-02-28”. Add this as a straight horizontal line to your plot from above with the geom_hline() function

```{r}
avg.36 <- mean(vax.36$percent_of_population_fully_vaccinated)
```

```{r}
ljplot + 
  geom_hline(yintercept = avg.36, color = "red", linetype = 2)
```

>Q17. What is the 6 number summary (Min, 1st Qu., Median, Mean, 3rd Qu., and Max) of the “Percent of Population Fully Vaccinated” values for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2023-02-28”? 

```{r}
summary(vax.36$percent_of_population_fully_vaccinated)
```

>Q18. Using ggplot generate a histogram of this data.

```{r}
ggplot(vax.36) +
  aes(vax.36$percent_of_population_fully_vaccinated) +
  geom_histogram() +
  xlim(0,1) +
  labs(x = "Percent Vaccinated")
```

>Q19. Is the 92109 and 92040 ZIP code areas above or below the average value you calculated for all these above?

```{r}
filter(vax.36, zip_code_tabulation_area %in% c("92109", "92040"))
```

>Q20. Finally make a time course plot of vaccination progress for all areas in the full dataset with a age5_plus_population > 36144.

```{r}
vax.36.all <- filter(vax, age5_plus_population > 36144)

ggplot(vax.36.all) + 
  aes(x = as_of_date, y = percent_of_population_fully_vaccinated, group = zip_code_tabulation_area) + 
  geom_line(alpha = 0.2, color = "blue") +
  ylim(0,1) +
  labs(x = "Date", y = "Percent Vaccinated", title = "Vaccination rate across California", subtitle = "Areas with population >36k") +
  geom_hline(yintercept = avg.36, linetype = 2)
```
